---
title: "Reproducible Research Project-1"
author: "Raji"
date: "April 13, 2016"
output: html_document
---


###Loading and Preprocessing the Data to create tidy data set 

```{r}
library(lattice)

setwd("C:/Raji Data/Coursera DataScience/Reproducible Research/Week1/Project1/activity")

# Load the raw activity data
activity_raw <- read.csv("activity.csv", stringsAsFactors=FALSE)

# Show few rows from raw data
head(activity_raw)

```

###Transform the data into a format suitable for analysis

```{r}
# Transform the date attribute to an actual date format
activity_raw$date <- as.POSIXct(activity_raw$date, format="%Y-%m-%d")

# Get name of the  weekday from the date attribute using weekdays function and add to the raw data
activity_raw <- data.frame(date=activity_raw$date, 
                           weekday=tolower(weekdays(activity_raw$date)), 
                           steps=activity_raw$steps, 
                           interval=activity_raw$interval)

# Compute the day type (weekend or weekday) and add this column to raw data
activity_raw <- cbind(activity_raw, 
                      daytype=ifelse(activity_raw$weekday == "saturday" | 
                                       activity_raw$weekday == "sunday", "weekend", 
                                     "weekday"))

# Create the final data.frame - the tidy data set
tidy_activity <- data.frame(date=activity_raw$date, 
                       weekday=activity_raw$weekday, 
                       daytype=activity_raw$daytype, 
                       interval=activity_raw$interval,
                       steps=activity_raw$steps)


head(tidy_activity)

```

### What is the mean total number of steps taken per day?

For this part of the assignment,  the missing values in the dataset are ignored.
Make a histogram of the total number of steps taken each day.
Compute the total number of steps each day with NA values removed

```{r}
totalsteps_by_date <- aggregate(tidy_activity$steps, by=list(tidy_activity$date), FUN=sum, na.rm=TRUE)

## Rename the attributes
names(totalsteps_by_date) <- c("Date", "Total_Steps")

## Show few rows  from totalsteps_by_data 
  head(totalsteps_by_date)
```

### Histogram of Total Number of Steps For Each day:

```{r, echo=FALSE}
 hist(totalsteps_by_date$Total_Steps,
       breaks=seq(from=0, to=25000, by=2500), 
       col="red", 
       main="Total Number of Steps Taken Each Day",
       xlab="Date",
       ylab="Total Steps") 
```

### Calculate and report the mean and median total number of steps taken per day
The mean and median are calculated as below
```{r}
  print(mean(totalsteps_by_date$Total_Steps)) 
  print(median(totalsteps_by_date$Total_Steps))
  
```


###What is the Average Daily Activity Pattern?
```{r}
avg_daily_activity <- aggregate(tidy_activity$steps, 
                         by=list(tidy_activity$interval), 
                         FUN=mean, 
                         na.rm=TRUE)

names(avg_daily_activity) <- c("interval", "average")
  
```
### Time Series Plot to Show Average Daily Activity
```{r, echo=FALSE}
plot(avg_daily_activity$interval, 
     avg_daily_activity$average, 
     type="l", 
     col="green", 
     lwd=2, 
     xlab="Interval [minutes]", 
     ylab="Average Number of Steps", 
     main="Average number of steps per time interval")
```

###Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
# Find the position of the maximum mean
max_pos <- which(avg_daily_activity$average == max(avg_daily_activity$average ))

# We lookup the value of interval at this position
max_interval <- avg_daily_activity[max_pos, 1]
print(max_interval)  ## 835

```
### Imputing missing values
```{r}
#Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

print(NA_count <- sum(is.na(tidy_activity$steps)))  ## 2304

#Create a new dataset that is equal to the original dataset but with the missing data filled in with the mean of steps
new_dataset <- tidy_activity

new_dataset[is.na(tidy_activity$steps), ]$steps <- mean(activity_raw$steps, na.rm = TRUE)
print(head(new_dataset))

```
### Compute Total number of steps taken each day after NA substitution
```{r}
avg_data_after_NA_substitution <- aggregate(new_dataset$steps, by=list(new_dataset$date), FUN=sum)

# Rename the attributes
names(avg_data_after_NA_substitution) <- c("date", "total")

```
### Histogram of the total number of steps each day
```{r, echo=FALSE}
hist(avg_data_after_NA_substitution$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 30), 
     main="Total Number of Steps Taken Each Day\n(NA replaced by mean value)")
```

###Effect of NA substitution on mean and median total number of steps taken per day.

```{r}
print(mean(avg_data_after_NA_substitution$total))  
print(median(avg_data_after_NA_substitution$total)) 

```

Mean and Median have increased because NAs were substituted by average steps per day.


### Are there differences in activity patterns between weekdays and weekends?

Factor variable with two levels - "weekday" and "weekend"   is already in the tidy dataset

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r}
## Compute Mean
mean_data <- aggregate(new_dataset$steps, 
                       by=list(new_dataset$daytype, 
                               new_dataset$weekday, new_dataset$interval), mean)

## Rename the column names
names(mean_data) <- c("daytype", "weekday", "interval", "mean")

head(mean_data)

```
### Panel Plot Showing Difference in Activity Categorized by WeekDay and WeekEnd

```{r, echo=FALSE}

xyplot(mean ~ interval | daytype, mean_data, 
       type="l", 
       lwd=1, 
       xlab="Interval", 
       ylab="Number of steps", 
       layout=c(1,2))
```



